\Huge{Machine Learning} \\ \LARGE Nature's Quest: Deep Learning Exploration of Global Plant Traits from Images and Geodata

### 2024-03-26 paper/paper.tex
paper stuff

% 5-fold stratified cross-validation + verification and split

@@

>>>>

% 5-fold stratified cross-validation + verification and split 
% \cite{Marcot2021} k = 5 good for large sample
% \cite{kequal10isgood} k ~ 10 generally good choice
Another method of reducing overfitting we thought relevant to apply here would be k-fold cross validation. For the choice of $k$ we chose 5 as for large datasets this has proven effective \cite{Marcot2021} and is generally around the commonly chosen value of 10 \cite{kequal10isgood}. Given that the data is from a geographically large enough sample we decided to create the folds in a stratified manner under the assumption the distribution of traits would be reasonably representaitive of the true distribution and thus one the model should learn. This is important as it ensures that the model is trained on a representative sample of the data in each fold and in turn has its predictions distributed in a similar manner to the assumed true distribution of the traits. So we reduce the risk of overfitting to the training data while also ensuring the model is trained using a representaitive sample of targets in each fold. To validate that the distributions of the target traits are indeed similar we averaged the distributions of the target traits in each fold and overlayed this on the original distribution of trait means and normalized them to the same scale.

\begin{figure}[!h]
    \centering
    \includegraphics[width=1\textwidth]{assets/distribution_match_folds.png}
    \caption{Comparison of the average trait distributions in each fold with the original distribution of trait means ; (blue) normalized distribution of the trait means across the entire dataset ; (yellow) average normalized distribution of the trait means in each fold}
\end{figure}

@@

% give an overview of the model 
    % activation function 
    % batch size
    % optimizer 
    % epochs
Our image CNN is a multi-output model which takes in a tensor consisting of red, green, and blue channels with an image size of 224x224 pixels. The model consists of a pretrained backbone; in our instance we compared \textit{efficientnetv2\_b2\_imagenet}, \textit{efficientnetv2\_s\_imagenet}, and \textit{resnet50\_imagenet}. The backbone outputs then feed into a global average pooling layer, followed by 2 dense layers and a dropout layer connected to the two dense output layers for the main and auxillary task. The hidden layers use a ReLU activation function, the outputs use no activation function as the targets are continuous. The model is trained using the Adam optimizer with a learning rate of 1e-4, a batch size of 48 and for 12 epochs. Additionally we used the $R^2$ metric for both loss and evaluation of the model performance as this is a regression problem. Finally we used step-based for the learning rate schedule which is defined as follows
\[
    \text{lr} = \text{lr}_\text{max} \times d^{\floor*{\frac{e_d}{r}}}
\]
Where $e_d$ is the current epoch, $r$ is the step size (=2), and $d$ is the decay rate.

% talk about the backbones 
    % ResNet 
    % EfficientNet b2 
    % EfficientNet s
\smallskip
Residual Network (ResNet) \cite{he2016identity} introduced in 2016 is a deep learning model that is able to learn features from images. Most famously its known for mitigating the vanishing gradient problem through the introduction of identiy skip connections. We are using \text{ResNet50} which was trained on imagenet and consists of 48 convolution layers, in addition to a MaxPool and AveragePool layer. EfficientNetV2 \cite{tan2021efficientnetv2} is a family of CNNs from 2021 that have faster speeds than many previous models primarily by increasing image size during training and incorporating adaptive regularization to compensate for any potential accuracy loss. We are using \textit{efficientnetv2\_b2\_imagenet} and \textit{efficientnetv2\_s\_imagenet} which are pretrained on imagenet, with the ladder being a smaller (fewer parameters) version of the former. Comparing these backbones in a table we can get some overview of how they stack up as seen in \ref{tab:backbone_comparison}

To compare these backbones we trained 3 different models with the same architecture but different backbones. We also specified a contribution of 0.3 to the final loss of the auxillary task (as opposed to 1.0 for the main task) as to limit its impacts on the direction of training. A visual overview of the different models can be seen in \ref{fig:models_overview}.

\begin{table}[!h]
    % ref for this table

    \centering
    \begin{tabular}{@{}llll@{}}
    \toprule
    & EfficientNetV2 b2 & EfficientNetV2 s & ResNet50 \\ \midrule
    parameters              & 8.77M             & 20.33M           & 23.56M   \\
    imagenet top 5 accuracy & 94.9\%            & 96.7\%           & (missing)       \\ \bottomrule
\end{tabular}
\caption{Comparison of the different backbones and published top 5 accuracy on imagenet for the specific models}
\label{tab:backbone_comparison}
\end{table}

\begin{figure}[!h]
    \centering
    \includegraphics[width=1\textwidth]{assets/different_models.png}
    \caption{Visual overview of the 3 models with different backbones}
    \label{fig:models_overview}
\end{figure}
